---
title: "Tornadoes Analysis"
author: "Abigail Castro, Maggie Salomonsky, Aditi Jain, Sarah Saas, Ethan Thurmond"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Note: Divide the task among yourself. Each team member should contribute to at least one part of the assignment. The person whose name starts last in the alphabetic order shares screen and compiles the Rmd file. 

# Tidy Tornadoes

The US Storm Prediction Center make severe weather data available from the website <http://www.spc.noaa.gov/wcm/#data>. This data is used by insurance companies to help with their claims evaluation and forecasting. A description of the data can be found <http://www.spc.noaa.gov/wcm/data/SPC_severe_database_description.pdf>.


1.  Download the [tornado event data](https://drive.google.com/file/d/1x0T23hAyGZfBvMFw-nbqwj01VnT9snFY/view?usp=sharing>) and import it in R
```{r}
library(tidyverse)
tornado <- read.csv(file = 'tornado.csv')
```


2.  Calculate the number of tornadoes by *year* and *Fujita score* (`f`) 
```{r}
d1<- tornado %>%  group_by(yr, f) %>%
     summarise(count = n())
d1

```


3.  Convert the results to a table (use pivot_wider). Note: Some years have 0 EF-5 tornadoes. The final result should look like this

|   year| EF-0| EF-1| EF-2| EF-3| EF-4| EF-5|
|----:|----:|----:|----:|----:|----:|----:|
| 2007|  681|  306|   97|   27|    4|    1|
| 2008|  997|  515|  158|   56|   11|    1|
| 2009|  709|  355|   94|   21|    3|    0|
| 2010|  776|  351|  129|   42|   17|    0|
| 2011|  821|  638|  212|   72|   25|    9|
| 2012|  577|  242|  100|   32|    5|    0|
| 2013|  508|  314|   86|   22|    8|    1|
| 2014|  478|  325|   76|   20|    7|    0|
| 2015|  704|  415|   69|   19|    5|    0|   

```{r}
d1 <- d1 %>%
    pivot_wider(names_from = f, values_from = count)
d1[is.na(d1)] <- 0
d1
```



4. What is the type of EF variables? How do you determine that?
```{r}
d1
```
By using the above command, you can see that the EF variables are of type int.

5. Compare different types of tornadoes in terms of 1) frequency of occurrence and 2) trends over the years. What type of plots do you need to use? How do you determine that? Look at the data types to see if you need to convert the variable types.
Since the year is a categorical variable and count is an int, we should use bar charts to visualize the data.
There is no need to convert the variable types.
```{r}
d2 <- tornado %>%  group_by(yr, f) %>%
     summarise(count = n())
ggplot(d2, aes(yr, count , fill= f)) +
   geom_bar(stat="identity", position = "dodge") +
   labs(x= 'Year', y= 'Number of Tornadoes')
ggplot(d2, aes(yr, count)) +
   geom_bar(stat="identity", position = "dodge") +
   labs(x= 'Year', y= 'Number of Tornadoes')
```


6. Describe your observations.
Generally, tornado frequency has decreased since 2008. 
There is no apparent correlation between severity of tornados and year, however, within each year there is a negative correlation between severity of tornados and frequency of tornados. 


# Time of occurance
The `time` column in the `tornado` data gives the time-of-day (24 hour clock, central time zone) when the tornado occurred. Ignoring the time zone issue, create a density plot of the fractional hour when tornadoes occur. 

1. Use the `separate()` function to create three new columns (*hour*, *min*, *sec*) from the `time` column. 
```{r}
tornado_data_sep <- tornado %>% 
  separate(time, into = c("hour", "min", "sec"), convert = TRUE)
```


2. Add another column, named `time2`, that gives the fractional number of hours that a tornado occurred. 
```{r}
tornado_data_sep <- tornado_data_sep %>% mutate(time2 = hour/24)

```

    
3. Generate a density plot of `time2`. Are there any differences by severity? 
```{r}
ggplot(tornado_data_sep, aes(x=time2)) + geom_density()
```


4. Describe your observations. 
As shown in the density plot there are certain hours of the day where tornados are more likely to occur.



